Lifting CDCL to Relational Domains

### 2016-10-15 domains.tex

%-------------------------------------------------------------------------------
\subsection{Properties of Abstract Domains}
%-------------------------------------------------------------------------------
%
\Omit {
An important property of a clause learning SAT solver is that each
element of the partial assignments domain is decomposable into
complementable elements.  This property of the domain is required to
learn elements that help to guide the model search away from the
conflicting region of the search space.
%
Most numerical abstract domains, such as intervals and octagons lack
complements, but they can be represented as intersections of
complementable half-spaces, each of which has a complement. 
}

We identify specific properties of domain elements necessary for 
abstract model search and learning in abstractions.  We formalise 
this in the sequel.
%
\begin{definition} 
A \emph{meet irreducible} $m$ in a complete lattice 
structure $A$ is an element with the following property.
\[\forall m_1, m_2 \in A: m_1 \meet m_2 = m \implies (m = m_1 \lor m = m_2), m \neq \top \]  
\rmcmt{mention minimum complementable element}
\end{definition}
%
The meet irreducibles in boolean domain $\mathcal{B}$ are 
$\{x \mapsto \true\}$ and $\{x \mapsto \false\}$. The meet 
irreducibles in the template polyhedra domain are all elements 
that concretise to half-spaces, i.e.\ they can be represented 
by a single inequality. For the interval domain, these are 
\ $x \leq d$ or $x \geq d$ for constants $d$.

@@

\begin{definition} 
A meet irreducible $m\in A$ is called \emph{complementable} iff $\exists
\bar{m} \in A: m\meet\bar{m}=\bot \wedge m\join\bar{m}=\top$.  
%A complementable meet irreducible $\bar{M}$ of an abstract lattice $A$ is the complement of a meet 
%irreducible $M \in A$ such that $\bar{M} \in A$ and the concretisation of $M$ 
%is the set complement of $\bar{M}$.  
%If every meet irreducible in $A$ is complementable, then $A$ is said to
%be have complementable meet irreducibles.
\end{definition}

An important property of a clause learning SAT solver 
is that a partial assignments domain can be decomposed 
into a set of precisely complementable singleton assignments.  
This property of the partial assignments domain helps to 
guide the model search away from the conflicting region 
of the search space and allows case-based analysis.  Complementation 
operator in abstract domains is different from the notion of precise
complements. 
%
\begin{definition}{(Precise complement)} Let $A$ be an lattice. A precise 
complement of an abstract element $a \in A$ is an element $\bar{a} \in A$ 
such that the negation of concretization $(\gamma)$ of $a$ is equivalent to 
concretization of $\bar{a}$, that is, $\gamma(\bar{a}) = \neg \gamma(a))$.
\end{definition}
%
The precise complementation property of partial assignment lattice 
can be generalized to other lattice structures.  Most numerical 
abstract domains, such as intervals, octagons, polyhedra, can be 
decomposed into half-spaces, that admits precise complements.  
For example, intervals lack complements, but the meet decomposition 
of intervals into meet-irreducibles have precise complements.  
Numerical abstract domains that admit complementable decomposition 
are shown in Figure~\ref{complement}.  Our template-based domain 
implementation precisely complements a meet irreducible in an abstraction.
% 
\begin{table}
\begin{center}
{
\begin{tabular}{l|l}
\hline
Domain & Precise Complements \\ \hline 
Interval & \((x \leq n) \quad \longrightarrow (x > n)\) \\ \hline
Octagon & \((x+y < 1) \quad \longrightarrow (-x-y < 0)\) \\ \hline
Equality & \((x==y) \quad \longrightarrow (x \neq y)\) \\ \hline 
\end{tabular}
}
\end{center}
\label{fig:complement}
\end{table}

\Omit {
The meet irreducibles in the partial Boolean assignments domain and
template polyhedra domains are all complementable.
%
Note that for template polyhedra, all non-meet-irreducible elements $e$
(except $\bot$ and $\top$) are not complementable within $A$.
%
However, we can complement each element in the meet decomposition of $e$ and
re-interpret the obtained set as a disjunction. 
%
Thus the complement of the octagon in Figure~\ref{octagon} 
can be written as a disjunction of meet irreducibles:
\[(x\leq -3) \lor (x\geq 2) \lor (y\leq -2) \lor (y\geq 3) \lor (x+y\geq 3) \lor (x-y\geq 2) \lor (y-x\geq 4) \lor (-x-y\geq 3)\]

Standard abstract interpretation does not require complementation property.  
Abstract Domain library, such as APRON C library~\cite{apron} is specialized 
for abstract interpretation.  We implement a template-based polyhedra domain 
that returns precise complements by decomposing it into half-spaces.    
}

>>>>

%-------------------------------------------------------------------------------
\subsection{Properties of Abstract Domains}
%-------------------------------------------------------------------------------
%
An important property of a clause learning SAT solver is that each
element of the boolean domain is decomposable into
complementable elements.  This property of the domain is required to
learn elements that help to guide the model search away from the
conflicting region of the search space.
%
Most numerical abstract domains, such as intervals and octagons lack
complements, but they can be represented as intersections of
complementable half-spaces, each of which has a complement. 
%
% We identify specific properties of domain elements necessary for 
% abstract model search and learning in abstractions.  
We formalise this in the sequel.
%
\begin{definition} 
A \emph{meet irreducible} $m$ in a complete lattice 
structure $A$ is an element with the following property.
\[\forall m_1, m_2 \in A: m_1 \meet m_2 = m \implies (m = m_1 \lor m = m_2), m \neq \top \]  
%\rmcmt{mention minimum complementable element}
\end{definition}
%
The meet irreducibles in the boolean domain $\mathcal{B}$ for a variable $x$ are 
$x$ and $\neg x$. The meet 
irreducibles in the template polyhedra domain are all elements 
that concretise to half-spaces, i.e.\ they can be represented 
by a single inequality. For the interval domain, these are 
\ $x \leq d$ or $x \geq d$ for constants $d$.

@@

\begin{definition} 
An element $a\in A$ is called \emph{precisely complementable}
iff $\exists \bar{a} \in A: \neg\gamma(\bar{a})=\gamma(a)$, 
i.e.\ there is an element whose negated concretisation equals
the concretisation of $a$.
%m\meet\bar{m}=\bot \wedge m\join\bar{m}=\top$.
%A complementable meet irreducible $\bar{M}$ of an abstract lattice $A$ is the complement of a meet 
%irreducible $M \in A$ such that $\bar{M} \in A$ and the concretisation of $M$ 
%is the set complement of $\bar{M}$.  
%If every meet irreducible in $A$ is complementable, then $A$ is said to
%be have complementable meet irreducibles.
\end{definition}

% An important property of a clause learning SAT solver 
% is that a partial assignments domain can be decomposed 
% into a set of precisely complementable singleton assignments.  
% This property of the partial assignments domain helps to 
% guide the model search away from the conflicting region 
% of the search space and allows case-based analysis.  Complementation 
% operator in abstract domains is different from the notion of precise
% complements. 
%
% \begin{definition}{(Precise complement)} Let $A$ be an lattice. A precise 
% complement of an abstract element $a \in A$ is an element $\bar{a} \in A$ 
% such that the negation of concretization $(\gamma)$ of $a$ is equivalent to 
% concretization of $\bar{a}$, that is, $\gamma(\bar{a}) = \neg \gamma(a))$.
% \end{definition}
%
% The precise complementation property of partial assignment lattice 
% can be generalized to other lattice structures.  Most numerical 
% abstract domains, such as intervals, octagons, polyhedra, can be 
% decomposed into half-spaces, that admits precise complements.  
% Numerical abstract domains that admit complementable decomposition 
% are shown in Figure~\ref{complement}.  Our template-based domain 
% implementation precisely complements a meet irreducible in an abstraction.
% % 
% \begin{table}
% \begin{center}
% {
% \begin{tabular}{l|l}
% \hline
% Domain & Precise Complements \\ \hline 
% Interval & \((x \leq n) \quad \longrightarrow (x > n)\) \\ \hline
% Octagon & \((x+y < 1) \quad \longrightarrow (-x-y < 0)\) \\ \hline
% Equality & \((x==y) \quad \longrightarrow (x \neq y)\) \\ \hline 
% \end{tabular}
% }
% \end{center}
% \label{fig:complement}
% \end{table}
% }

Note that for many domains $A$ including template polyhedra
most domain elements are not precisely complementable within $A$.
%
In fact, for template polyhedra all non-meet-irreducible elements $e$
(except $\bot$ and $\top$) are not precisely complementable,
whereas all meet irreducibles are precisely complementable.
%
Hence, we can complement each element in the meet decomposition of $e$ and
re-interpret the obtained set as a disjunction. 
%
Thus the complement of the octagon in Figure~\ref{octagon} 
can be written as a disjunction of meet irreducibles:
\[(x{\leq} -3) \lor (x{\geq} 2) \lor (y{\leq} -2) \lor (y{\geq} 3) \lor (x+y{\geq} 3) \lor (x-y{\geq} 2) \lor (y-x{\geq} 4) \lor (-x-y{\geq} 3)\]

Standard abstract interpretation does not require a complementation
operator. Hence, abstract domain libraries, such as APRON~\cite{apron}
do not provide it.  However, it can be implemented with the help of a
meet decomposition as explained above.
%
Another feature that abstract domain libraries do not provide is to
track the reasons why certain deductions have been made. We come back
to this point in Section~\ref{sec:abst}.

### 2016-10-15 graphs.tex

\section{Experiments}
\label{section:experiments}
\Omit {
We performed a number of experiments to demonstrate the utility and
applicability of {\algorithmName}.  All experiments were performed
on an Intel Xeon X5667 at 3\,GHz running Fedora 20 with 64-bit binaries.
Each individual run was limited to 13\,GB 
of memory and 900 seconds of
CPU time, % (with one exception noted below), 
enforced by the operating
system kernel.  We took the \emph{loops} meta-category (143 benchmarks) from the SV-COMP'15 benchmark
set.%
\footnote{\url{http://sv-comp.sosy-lab.org/2015/benchmarks.php}}
%

\subsection{\algorithmName\ Verifies More Programs Than the Algorithms it Simulates}

Table~\ref{tab:results} gives a comparison between \systemName\ running
\algorithmName\ (column 6) and \emph{the same system} running as an incremental
bounded model checker (IBMC) (column 2), incremental $k$-induction (i.e. without invariant inference, column 3)
and as an abstract interpreter (AI) (column 4).
%
 \algorithmName\ is more complete than each of the restricted modes. 
This is not self-evident since it could be much less efficient and,
thus, fail to solve the problems within the given time or memory limits.
%
$k$-induction can solve 60.8\% of the benchmarks, 13 more than IBMC. 
%
32\% of the benchmarks can be solved by abstract interpretation (bugs
are only exposed if they are reachable with 0 loop unwindings).  
%
\algorithmName\ solves 62.9\% of the benchmarks, 
proving 3 more properties than $k$-induction.

%%%%%%%%%%%%%%%%%%%%%%%%%% results table %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{table}
\centering
\begin{tabular}{|l|ccc|c|c|c|c|}
\hline
& IBMC & $k$-induction & AI & portfolio & algorithmName &
                                                               CPAchecker & ESBMC \\
\hline
counterexamples & \bf 38 & \bf 38 &  17 & \bf 38 & \bf 38 & 36 &  35 \\
proofs          &    36 &  49 &  30 &  51 &  52 & 59 & \bf 91 \\
false proofs    &     0 &   0 &   0 &   0 &   0 &  2 &  12\\
false alarms    &     2 &   2 &   0 &   2 &   2 &  2 &   0\\
inconclusive    &     0 &   0 &  93 &   0 &   0 &  4 &   2\\
timeout         &    65 &  53 &   3 &  50 &  51 & 38 &   2 \\
memory out      &     2 &   1 &   0 &   2 &   0 &  2 &   1 \\
total runtime   & 17.1h & 13.8h & 0.89h & 13.3h & 13.2h & 10.9h & 0.54h \\
\hline
\end{tabular}~\\[1ex]
\caption{Comparison between algorithmName, the algorithms it subsumes,
the portfolio, and CPAchecker. The rows false alarms and false proofs 
indicate soundness bugs of the tool implementations.
\label{tab:results}}
\end{table}

\subsection{\algorithmName\ is at Least as Good as Their Na\"ive Portfolio}

To show
that \algorithmName\ is more than a mixture of three techniques and
that they strengthen each other, consider column 5 of
Table~\ref{tab:results}.  This gives the results of an ideal portfolio
in which the three restricted techniques are run in parallel on and
the portfolio terminates when the first returns a conclusive result.
Thus the CPU time taken is three times the time taken by the fastest
technique for each benchmark (in practice these could be run in
parallel, giving a lower \emph{wall clock} time).
%
In our setup, \algorithmName\ had a disadvantage as each component of
virtual portfolio had the same %time and 
memory restriction as
{\algorithmName}, thus effectively giving the portfolio three times as much memory.

Still, {\algorithmName} is slightly faster and more
accurate than the portfolio as can be seen in Table \ref{tab:results}.
%
The scatter plot in Figure~\ref{fig:results}a shows the results for
each benchmark:
%
one can observe that {\algorithmName} is up to one order of magnitude
slower on many unsafe benchmarks, which is obviously due to the
additional work of invariant inference that
{\algorithmName} has to perform in contrast to IBMC.
%
However, note that {\algorithmName} is faster than the 
portfolio on some safe and even one unsafe benchmarks.
This suggests that {\algorithmName} is more than the sum of its parts.

\subsection{\algorithmName\ is Comparable with State-of-the-Art Approaches}

We compared our implementation of \algorithmName\ with
CPAchecker%
\footnote{SVCOMP'15 version, http://cpachecker.sosy-lab.org/}%
%
and ESBMC%
\footnote{SVCOMP'15 version, http://www.esbmc.org/}%
, which uses $k$-induction.
%
The results are shown in the last three columns in Table \ref{tab:results}
and in the scatter plot in Figure~\ref{fig:results}b. Additional
results are given in \pponly{the extended version
  \cite{extended-version}}\rronly{Appendix \ref{sec:further}}.
%
In comparison to CPAchecker, the winner of SVCOMP'15,
our prototype of {\algorithmName} is overall a bit slower and proves 
fewer properties (due to more timeouts), but as
Figure~\ref{fig:results}b shows, it significantly outperforms
CPAchecker on most benchmarks.
%
ESBMC exposes fewer bugs, but proves many more properties and is 
significantly faster. However, it has 6 times
more soundness bugs than our implementation.%
\footnote{The two false alarms in our current implementation are due to 
limited support for dynamic memory allocation.}
%
These results show that our prototype implementation of
\algorithmName\ can keep up with state-of-the-art verification tools.

}
%%%%%%%%%%%%%%%%%%%%%%%%%% kiki vs vbs vs CPAchecker %%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[t]
\begin{tabular}{@{\hspace{-1.5em}}c@{\hspace{1em}}c}
\begin{tikzpicture}[scale=0.75]
	\begin{loglogaxis} [xmin=.1,xmax=1500, ymin=.1, ymax=1500, xlabel=kIkI (time in seconds),
			ylabel=portfolio (time in seconds), 
			legend style={at={(0.8,0.15)},
			anchor=north,legend columns=-1 },
			]
\addplot [mark size=1pt,only marks,scatter,point meta=explicit symbolic,
	scatter/classes={s={mark=square},u={mark=triangle*,blue}},] 
	table [meta=label] {scatter-kiki-vbs-sas.dat};
	\legend{Safe,Unsafe}
\addplot [domain=.1:1500] {x};
\addplot [red,sharp plot, domain=.1:1500] {900}
          node [below] at (axis cs:10,850) {timeout};
\addplot [red,sharp plot, domain=.1:1500] coordinates{(900,.1) (900,1500)}
          node [left,rotate=90] at (axis cs:700,10) {timeout}
 node [right,black] at (axis cs:10,3) {portfolio faster}
 node [right,black] at (axis cs:1,55) {kIkI faster};
%\addplot [red,sharp plot, update limits=false] coordinates{(900,.1) (900,1500)}
%	node [left] at {axis cs:700,200} {timeout};
\end{loglogaxis}
\end{tikzpicture}
 &
\begin{tikzpicture}[scale=0.75]
	\begin{loglogaxis} [xmin=.1,xmax=1500, ymin=.1, ymax=1500, xlabel=kIkI (time in seconds),
			ylabel=CPAchecker (time in seconds),
			legend style={at={(0.8,0.15)},
			anchor=north,legend columns=-1 },
			]
\addplot [mark size=1pt,only marks,scatter,point meta=explicit symbolic,
	scatter/classes={s={mark=square},u={mark=triangle*,blue}},] 
	table [meta=label] {scatter-kiki-vbs-sas.dat};
	\legend{Safe,Unsafe}
\addplot [domain=.1:1500] {x};
\addplot [red,sharp plot, domain=.1:1500] {900}
          node [below] at (axis cs:10,850) {timeout};
\addplot [red,sharp plot, domain=.1:1500] coordinates{(900,.1) (900,1500)}
          node [left,rotate=90] at (axis cs:700,150) {timeout}
 node [right,black] at (axis cs:10,3) {CPAchecker faster}
 node [right,black] at (axis cs:1,55) {kIkI faster};
\end{loglogaxis}
\end{tikzpicture} \\
(a) & (b)
\end{tabular}
\caption{\label{fig:results}
Runtime Comparison
}
\end{figure}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%% scatter plots %%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[t]
\begin{tabular}{@{\hspace{-1.5em}}c@{\hspace{1em}}c}
\begin{tikzpicture}[scale=0.75]
	\begin{loglogaxis} [xmin=.1,xmax=1500, ymin=.1, ymax=1500, xlabel=kIkI (time in seconds),
			ylabel=portfolio (time in seconds), 
			legend style={at={(0.8,0.15)},
			anchor=north,legend columns=-1 },
			]
\addplot [mark size=1pt,only marks,scatter,point meta=explicit symbolic,
	scatter/classes={s={mark=square},u={mark=triangle*,blue}},] 
	table [meta=label] {plotdata/scatter-plot1.dat};
	\legend{Safe,Unsafe}
\addplot [domain=.1:1500] {x};
\addplot [red,sharp plot, domain=.1:1500] {900}
          node [below] at (axis cs:10,850) {timeout};
\addplot [red,sharp plot, domain=.1:1500] coordinates{(900,.1) (900,1500)}
          node [left,rotate=90] at (axis cs:700,10) {timeout}
 node [right,black] at (axis cs:10,3) {portfolio faster}
 node [right,black] at (axis cs:1,55) {kIkI faster};
%\addplot [red,sharp plot, update limits=false] coordinates{(900,.1) (900,1500)}
%	node [left] at {axis cs:700,200} {timeout};
\end{loglogaxis}
\end{tikzpicture}
 &
\begin{tikzpicture}[scale=0.75]
	\begin{loglogaxis} [xmin=.1,xmax=1500, ymin=.1, ymax=1500, xlabel=kIkI (time in seconds),
			ylabel=CPAchecker (time in seconds),
			legend style={at={(0.8,0.15)},
			anchor=north,legend columns=-1 },
			]
\addplot [mark size=1pt,only marks,scatter,point meta=explicit symbolic,
	scatter/classes={s={mark=square},u={mark=triangle*,blue}},] 
	table [meta=label] {plotdata/scatter-plot1.dat};
	\legend{Safe,Unsafe}
\addplot [domain=.1:1500] {x};
\addplot [red,sharp plot, domain=.1:1500] {900}
          node [below] at (axis cs:10,850) {timeout};
\addplot [red,sharp plot, domain=.1:1500] coordinates{(900,.1) (900,1500)}
          node [left,rotate=90] at (axis cs:700,150) {timeout}
 node [right,black] at (axis cs:10,3) {CPAchecker faster}
 node [right,black] at (axis cs:1,55) {kIkI faster};
\end{loglogaxis}
\end{tikzpicture} \\
(a) & (b)
\end{tabular}
\caption{\label{fig:results}
Comparison
}
\end{figure}
